\red{Cluster-efficient} pangenome graph construction with nf-core/pangenome

### 2024-09-10 manuscript/fig_panel.tex
hart aber herzlich
we have a 4 pager :)
other reviewer comments

\begin{figure*}[htpb]
	\sidesubfloat[First subfigure]{%
	\includegraphics[width=\textwidth]{fig/pangenome_workflow.pdf}
	\label{fig:workflow}
	}
	\hfill
	\sidesubfloat[Second subfigure]{%
	\includegraphics[width=0.45\textwidth]{fig/chr19_histgrowth.pdf}
	\label{fig:hist_chr19}
	}
	\sidesubfloat[Second subfigure]{%
	\includegraphics[width=0.45\textwidth]{fig/ecoli2013_histgrowth.pdf}
	\label{fig:hist_ecoli}
	}
	%\sidesubfloat[Second subfigure]{%
	%\includegraphics[width=0.3\textwidth]{fig/chr19_histgrowth.pdf}
	%\label{fig:hist_ecoli}
	%}
	\caption{\textbf{(a)} Schematic representation of the nf-core/pangenome workflow processes and detailed analysis steps. The input consists of one FASTA file containing all sequences. The pipeline comes with 3 major entry points: Community detection (1), alignment distribution (2), and core workflow (3). Optional community detection (1) is performed on the input sequences. If selected, the heavy all-to-all baise-pair level alignments (2) can be split into problems of equal size. nf-core/pangenome’s core workflow (3) is a direct mirror of PGGB. If running in community mode, all communal graphs are combined into one (4) and the subsequent quality control subworkflow is executed. The output is a pangenome graph in GFA format. \textbf{(b) + (c)} Pangenome growth curves of the built pangenome graphs. Growth type is defined as the minimum fraction of haplotypes that must share a graph feature after each time a haplotype is added to the growth histograph. $quorum>=0$: All sequences without any filtering are considered. $quorum>=10$: Sequences traversed by at least 10\% of the haplotypes. $quorum>=50$: Sequences traversed by at least 50\% of haplotypes. $quorum>=95$: Sequences traversed by 95\% of haplotypes. \textbf{(b)} Pangenome growth curve of the chromosome 19 pangenome graph of 1000 haplotypes. \textbf{(c)} Pangenome growth curve of the E. coli pangenome graph of 2013 haplotypes.
	}
	\label{fig:panel}
\end{figure*}

>>>>

\begin{figure*}[]
	\sidesubfloat[First subfigure]{%
	\includegraphics[width=\textwidth]{fig/pangenome_workflow.pdf}
	\label{fig:workflow}
	}
	\vspace{-0.3cm}
	\hfill
	\sidesubfloat[Second subfigure]{%
	\includegraphics[width=0.45\textwidth]{fig/chr19_histgrowth.pdf}
	\label{fig:hist_chr19}
	}
	\sidesubfloat[Second subfigure]{%
	\includegraphics[width=0.45\textwidth]{fig/ecoli2013_histgrowth.pdf}
	\label{fig:hist_ecoli}
	}
	%\sidesubfloat[Second subfigure]{%
	%\includegraphics[width=0.3\textwidth]{fig/chr19_histgrowth.pdf}
	%\label{fig:hist_ecoli}
	%}
	\vspace{-0.25cm}
	\caption{\textbf{(a)} Schematic representation of the nf-core/pangenome workflow processes and detailed analysis steps. The input consists of one FASTA file containing all sequences. The pipeline comes with 3 major entry points: \red{(1) Community detection, which identifies clusters of related sequences or regions in the pangenome graph to reveal biologically significant patterns like conserved or divergent areas across genomes~(Supplementary \ref{community}), (2) alignment distribution, and (3)  core workflow.} Optional community detection (1) is performed on the input sequences. If selected, the heavy all-to-all baise-pair level alignments (2) can be split into problems of equal size. nf-core/pangenome’s core workflow (3) is a direct mirror of PGGB. If running in community mode, all communal graphs are combined into one (4) and the subsequent quality control subworkflow is executed. The output is a pangenome graph in GFA format. \textbf{(b) + (c)} Pangenome growth curves of the built pangenome graphs. Growth type is defined as the minimum fraction of haplotypes that must share a graph feature after each time a haplotype is added to the growth histograph. $quorum>=0$: All sequences without any filtering are considered. $quorum>=10$: Sequences traversed by at least 10\% of the haplotypes. $quorum>=50$: Sequences traversed by at least 50\% of haplotypes. $quorum>=95$: Sequences traversed by 95\% of haplotypes. \textbf{(b)} Pangenome growth curve of the chromosome 19 pangenome graph of 1000 haplotypes. \textbf{(c)} Pangenome growth curve of the \textit{E. coli} pangenome graph of 2013 haplotypes.
	}
	\label{fig:panel}
\end{figure*}
\vspace{-0.4cm}

### 2024-09-10 manuscript/main.tex
other reviewer comments

	The first step in the nf-core/pangenome pipeline is the all-to-all alignment of the input sequences with the whole-chromosome pairwise sequence aligner WFMASH \citep{wfmash}. 
	This avoids reference, order, or orientation bias, and allows each sequence in the pangenome to serve as a reference when exploring related variation. 
	In the pangenome graph induction step SEQWISH \citep{Garrison2022}, an alignment to variation graph inducer, converts the sequence alignments into a variation graph. 
	%SEQWISH can recapture homology relationships that may not be encoded in the initial alignments. 
	%This allows us to randomly sparsify the mappings computed by WFMASH. 
	%The applied heuristic is based on the Erdös-Rényi random graph model \citep{Bollobs2001}. 
	We then normalize the graph with the variation graph simplification algorithm SMOOTHXG \citep{Garrison2023}:
	A 1-dimensional (1D) graph embedding \citep{Heumos2023} orders the graphs’ nodes to best-match the nucleotide distances of the genomic paths of the graph.
	%This considers the local spurious complexity inherent in the raw pairwise alignments. 
	%If ignored, this could lead to difficulties in downstream processing of the graph.
	%Therefore, SMOOTHXG iteratively applies a local Multiple Sequence Alignment (MSA), partial order alignment (POA) \citep{Lee2002}, to refine and compress the pangenome graph. A 1-dimensional (1D) graph embedding \citep{Heumos2023} gives rise to the regions the kernel is applied to. 
	%After applying the graph embedding algorithm, the graphs’ node order best-matches the nucleotide distances of the genomic paths of the graph. 
	Next, the graph is split into partially overlapping segments. 
	%The sequences of each segment are realigned with POA. 
	The sequences of each segment are realigned with a local Multiple Sequence Alignment (MSA) kernel, partial order alignment (POA) \citep{Lee2002}.
	Afterwards, the segments are laced back together into a variation graph. 
	By default, the SMOOTHXG process is applied 3 times in order to smoothen the edge effects at the boundaries of the segments.
	Finally, we employ GFAFFIX \citep{Liao2023} to systematically condense redundant nodes within the graph.
	%In the final normalization phase, we employ GFAFFIX \citep{Liao2023} to systematically condense redundant nodes within the graph. 
	%Subsequently, the 1D graph embedding algorithm is reapplied to ensure a human-readable representation of the processed data. 
	%For more details about each step, we refer to the original PGGB paper (Garrison et al. 2023).

	Basic graph build quality is evaluated with ODGI: Optimized Dynamic Genome/Graph Implementation \citep{Guarracino2022} for understanding pangenome graphs. ODGI reports basic graph statistics and diagnostic 1D and 2D visualizations. 
	%Specifically, ODGI reports basic graph statistics, such as the length, and the number of nucleotides, paths, edges, nodes, and components. ODGI’s human readable 1D and 2D visualizations help to understand if the biology encoded in the graph fits the expectations. 
	%Six different 1D views focus on displaying the relative alignments of the graph’s paths, while the 2D view highlights high-level features such as SVs. 
	%Additionally, the 2D TSV layout of ODGI can be interactively explored in the variation graph visualizers gfaestus (\href{https://github.com/chfi/gfaestus}{https://github.com/chfi/gfaestus}) or waragraph (\href{https://github.com/chfi/waragraph}{https://github.com/chfi/waragraph}). 
	Optionally, nf-core/pangenome calls variants against any (reference) path(s) in the graph using \textit{vg deconstruct} \citep{Garrison:2018}. 
	Finally, graph statistics and visualizations are summarized in a MultiQC \citep{Ewels_2016} report. Pipeline implementation details are given in Suppl.~\ref{implementation}.
	\vspace{-0.6cm}
	\section{Results}

@@

	We implemented nf-core/pangenome, an easy-to-install, portable, and cluster-scalable pipeline for the unbiased construction of pangenome variation graphs. 
	It is the first pangenomic nf-core pipeline enabling the comparative analysis of gigabase-scale pangenome datasets. 
	The pipeline’s core workflow steps were already successfully applied to \textit{Neisseria meningitidis} \citep{Yang2023}, wild grapes \citep{Cochetel2023}, humans \citep{Guarracino2023, Liao2023}, grapevines \citep{Guo2024}, taurines \citep{Milia2024}, and rats \citep{Villani2024} underpinning the community effort to focus on a best-practice workflow to create reference-unbiased and sequence complete pangenome graphs. 
	The modular domain-specific language (DSL) 2 pipeline structure eases the exchange of key processes with alternative tools, the extent of the pipeline with new tools, and the integration of parts of the pipeline with other (sub-)workflows.

@@

	We anticipate the pipeline, or its parts, will enhance current single linear reference analysis methods to explore whole population variation instead of focusing on one reference only. Looking ahead, pangenome construction pipelines like nf-core/pangenome will play a pivotal role in studying entire populations, single-cell whole genome sequencing analysis, and constructing personalized (medical) pangenome references \citep{Sirn2023}.
	%We believe that the pipeline or parts of it will enhance existing single linear reference analysis methods to explore whole population variation instead of focusing on one reference only. We envision for the future that pangenome construction pipelines, like nf-core/pangenome, will play a pivotal role in studying whole populations, single-cell whole genome sequencing analysis, and, for example, specifically when it comes to the construction of personalized (medical) pangenome references \citep{Sirn2023}.

	\vspace{-0.5cm}

	\section*{Software and data availability}

@@

	\subsection{Chromosome community detection}

	Eukaryotic genomes are usually organized into chromosomes. 
	Taking this into account during graph construction, the chromosome groupings from the input sequence are examined. 
	Specifically, the homologies detected in the all-to-all WFMASH mapping step are put into the Leiden \citep{Traag2019} clustering algorithm. 
	The edge weight is $mapped\_length * mapped\_identity$. 
	For each of the resulting communities, the nf-core/pangenome core workflow is executed in parallel. 
	The communal graphs are joined into one and a final round of quality control is applied (see Fig~\ref{fig:workflow}, brown tubes). 
	In practice, this works well for large input sequences with a large mapping length (>1Mb) filter, which was demonstrated by \cite{Guarracino2023} when exploring the recombination between heterologous human acrocentric chromosomes.

>>>>

	\red{The pipeline begins with an all-to-all alignment of the input sequences using the whole-chromosome pairwise sequence aligner WFMASH \citep{wfmash}. }
	This avoids reference, order, or orientation bias, and allows each sequence in the pangenome to serve as a reference when exploring related variation. 
	In the pangenome graph induction step SEQWISH \citep{Garrison2022}, an alignment to variation graph inducer, converts the sequence alignments into a variation graph. 
	%SEQWISH can recapture homology relationships that may not be encoded in the initial alignments. 
	%This allows us to randomly sparsify the mappings computed by WFMASH. 
	%The applied heuristic is based on the Erdös-Rényi random graph model \citep{Bollobs2001}. 
	\red{The graph is then simplified using SMOOTHXG \citep{Garrison2023}:}
	A 1-dimensional (1D) graph embedding \citep{Heumos2023} orders the graphs’ nodes to best-match the nucleotide distances of the genomic paths of the graph.
	%This considers the local spurious complexity inherent in the raw pairwise alignments. 
	%If ignored, this could lead to difficulties in downstream processing of the graph.
	%Therefore, SMOOTHXG iteratively applies a local Multiple Sequence Alignment (MSA), partial order alignment (POA) \citep{Lee2002}, to refine and compress the pangenome graph. A 1-dimensional (1D) graph embedding \citep{Heumos2023} gives rise to the regions the kernel is applied to. 
	%After applying the graph embedding algorithm, the graphs’ node order best-matches the nucleotide distances of the genomic paths of the graph. 
	Next, the graph is split into partially overlapping segments. 
	%The sequences of each segment are realigned with POA. 
	The sequences of each segment are realigned with a local Multiple Sequence Alignment (MSA) kernel, partial order alignment (POA) \citep{Lee2002}.
	Afterwards, the segments are laced back together into a variation graph. 
	By default, the SMOOTHXG process is applied 3 times in order to smoothen the edge effects at the boundaries of the segments.
	Finally, we employ GFAFFIX \citep{Liao2023} to systematically condense redundant nodes within the graph.
	%In the final normalization phase, we employ GFAFFIX \citep{Liao2023} to systematically condense redundant nodes within the graph. 
	%Subsequently, the 1D graph embedding algorithm is reapplied to ensure a human-readable representation of the processed data. 
	%For more details about each step, we refer to the original PGGB paper (Garrison et al. 2023).

	\red{Graph quality is assessed with ODGI~\citep{Guarracino2022}, which provides statistics and visualizations.} %: Optimized Dynamic Genome/Graph Implementation \citep{Guarracino2022} for understanding pangenome graphs. ODGI reports basic graph statistics and diagnostic 1D and 2D visualizations. 
	%Specifically, ODGI reports basic graph statistics, such as the length, and the number of nucleotides, paths, edges, nodes, and components. ODGI’s human readable 1D and 2D visualizations help to understand if the biology encoded in the graph fits the expectations. 
	%Six different 1D views focus on displaying the relative alignments of the graph’s paths, while the 2D view highlights high-level features such as SVs. 
	%Additionally, the 2D TSV layout of ODGI can be interactively explored in the variation graph visualizers gfaestus (\href{https://github.com/chfi/gfaestus}{https://github.com/chfi/gfaestus}) or waragraph (\href{https://github.com/chfi/waragraph}{https://github.com/chfi/waragraph}). 
	\red{Optionally, variants can be called against any (reference) path(s) in the graph using \textit{vg deconstruct} \citep{Garrison:2018}. 
	Results are summarized in a MultiQC \citep{Ewels_2016} report.} 
	Pipeline implementation details are given in Suppl.~\ref{implementation}.

\red{	If desired, the pipeline performs community detection to identify clusters of related sequences in the pangenome graph, revealing biological patterns such as conserved or divergent regions across genomes (Supplementary 5.2), with the core workflow executed for each community in parallel.}
	\vspace{-0.6cm}
	\section{Results}

@@

	We implemented nf-core/pangenome, an easy-to-install, portable, and cluster-scalable pipeline for the unbiased construction of pangenome variation graphs.  
    \red{It is the first pangenomic pipeline within the nf-core framework that enables the comparative analysis of gigabase-scale pangenome datasets. 
    While other tools such as Minigraph \citep{Li:2020} or PGR-TK \citep{Chin2023} also address pangenome analysis, nf-core/pangenome uniquely integrates this capability into the standardized nf-core framework, offering compatibility with a wide range of modular workflows and community-developed best practices.}

	The pipeline’s core workflow steps were already successfully applied to \textit{Neisseria meningitidis} \citep{Yang2023}, wild grapes \citep{Cochetel2023}, humans \citep{Guarracino2023, Liao2023}, grapevines \citep{Guo2024}, taurines \citep{Milia2024}, and rats \citep{Villani2024} underpinning the community effort to focus on a best-practice workflow to create reference-unbiased and sequence complete pangenome graphs. 
	The modular domain-specific language (DSL) 2 pipeline structure eases the exchange of key processes with alternative tools, the extent of the pipeline with new tools, and the integration of parts of the pipeline with other (sub-)workflows.

@@

	We anticipate the pipeline, or its parts, will enhance current single linear reference analysis methods to explore whole population variation instead of focusing on one reference only. Looking ahead, pangenome construction pipelines like nf-core/pangenome will play a pivotal role in studying entire populations, single-cell whole genome sequencing analysis, and constructing personalized (medical) pangenome references \citep{Sirn2023}.
	%We believe that the pipeline or parts of it will enhance existing single linear reference analysis methods to explore whole population variation instead of focusing on one reference only. We envision for the future that pangenome construction pipelines, like nf-core/pangenome, will play a pivotal role in studying whole populations, single-cell whole genome sequencing analysis, and, for example, specifically when it comes to the construction of personalized (medical) pangenome references \citep{Sirn2023}. 
	\vspace{-0.6cm}
	\section*{Software and data availability}

@@

	\subsection{Chromosome community detection}
	\label{community}

	\red{Chromosome community detection involves identifying clusters of closely related sequences or genomic regions within the pangenome graph. These communities represent areas of high similarity or shared sequence variants across multiple genomes. Detecting such communities allows researchers to focus on biologically meaningful patterns, such as conserved or highly divergent regions, which are crucial for understanding genetic diversity, evolutionary relationships, and genotype-phenotype associations.}

	\red{For example, eukaryotic genomes are typically organized into chromosomes, and this organization is considered during graph construction in community detection mode. Chromosome groupings from the input sequence are examined, and homologies detected in the all-to-all WFMASH mapping step are passed through the Leiden clustering algorithm \citep{Traag2019}, where the edge weight is calculated as $mapped\_length * mapped\_identity$. The nf-core/pangenome core workflow is then executed in parallel for each of the resulting communities. The communal graphs are subsequently merged into a single graph, followed by a final round of quality control (see Fig~\ref{fig:workflow}, brown tubes).}

	\red{This approach works effectively for large input sequences with a mapping length filter greater than 1 Mb, as demonstrated by \cite{Guarracino2023} when exploring recombination between heterologous human acrocentric chromosomes.}
